Higher-Order unification for free!

### 2024-04-10 paper/main.tex
update ref
clarify proofs

\newtheorem{invariant}{Invariant}
\crefname{invariant}{invariant}{invariants}

@@

\begin{lemma}
Properties \ref{prop:correct-ml} and
\ref{prop:complete-ml} hold for the implementation of \Uo above
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting \Ee is as strong as
\Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed) we can find this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the \Fo{} terms, the are two interesting cases:
\begin{itemize}
\item \elpiIn{fuva X ~\Uo~s}. In this case after \elpiIn{comp} we have
  $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
  $\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$.
\item \elpiIn{fapp[fuva X|L] ~\Uo~s}. In this case
 we have $Y_{\vec{x}} \Ue t$ that succeeds with
 $\sigma = \{ \vec{y} \vdash Y \mapsto t[\vec{x}/\vec{y}]\}$ that in turn
 is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{x}/\vec{y}]\}$.
 Thanks to \ref{clause:beta1}
 $(\lambda \vec{y}.s[\vec{x}/\vec{y}])~\vec{x} \Eo s$.
\end{itemize}
Since the mapping is a bijection occur check in \Ho{} corresponds to occur
check in \Fo{}.
\end{proof}

\begin{lemma} Properties simulation (\ref{prop:simulation}) and
fidelity (\ref{prop:fidelity}) hold
\end{lemma}
\begin{proof}[Proof sketch]
Since \elpiIn{progress1} is trivial \fstep and \hstep are the same, that is
in this
context where input terms are $\beta\eta$-normal and we disregard $\eta$-equivalence
\Ue is equivalent to \Uo.
\end{proof}

@@

@@

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
  \cref{prop:nf}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
  \cref{def:progressetaleft}, if 1) \lhs is a name, a constant of an
  application, then, $\lambda x.\lhs \appsep x$ is unified with \rhs. By
  \cref{inv:link-eta-right}, $\rhs = \lambda x. t$, therefore $\wellb(t)$.
  Otherwise, 2) \lhs has \elpiIn{lam} as functor, \rhs should not be an
  $\eta$-expansion, so, $\wellb(\rhs)$. In both cases, unification is
  performed between terms in \wellb.
\end{proof}

@@

\begin{proof}[Proof sketch]
  \lhs is variable, and, by \cref{def:progressetaright}, \rhs is either no more
  a \maybeeta, i.e. \rhs is not a $\eta$-expansion and, so, $\wellb(\rhs)$.
  Otherwise, \rhs can reduce to a term which cannot be a $\eta$-expansion, and,
  so, $\wellb(\rhs)$. In both cases, unification is done between terms in
  \wellb.
\end{proof}

@@

\begin{proof}
  The unification is done between the \rhs of two
  \linketa. Both \rhs has the shape $\lambda x.t$, with $\wellb(t)$, therefore,
  the unification is done between well-behaved terms.
\end{proof}

@@

\begin{proof}[Proof sketch]
  By \cref{lemma:unif-eta-1,lemma:unif-eta-2,lemma:unif-eta-3}, every
  unification performed by the activation of a \linketa is done between
  terms in \wellb, therefore, the substitution remains \wellb.
  \noteD[inline]{Bisogna aggiungere un lemma nella \cref{sec:nutshell} che dice che
  unificare due termini in \wellb, in una $\sigma$, tale che $\wellb(\sigma)$, 
  non invalida \wellb}
\end{proof}

>>>>

\newtheorem{invariant}{Invariant}
\crefname{invariant}{invariant}{invariants}
\crefname{equation}{property}{properties}

@@

\begin{lemma}
\Cref{prop:correct-ml,prop:complete-ml} hold for the implementation of \Uo above
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting \Ee is as strong as
\Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed) we can find this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the \Fo{} terms, the are two interesting cases:
\begin{itemize}
\item \elpiIn{fuva X ~\Uo~s}. In this case after \elpiIn{comp} we have
  $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
  $\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$.
\item \elpiIn{fapp[fuva X|L] ~\Uo~s}. In this case
 we have $Y_{\vec{x}} \Ue t$ that succeeds with
 $\sigma = \{ \vec{y} \vdash Y \mapsto t[\vec{x}/\vec{y}]\}$ that in turn
 is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{x}/\vec{y}]\}$.
 Thanks to \ref{clause:beta1}
 $(\lambda \vec{y}.s[\vec{x}/\vec{y}])~\vec{x} \Eo s$.
\end{itemize}
Since the mapping is a bijection occur check in \Ho{} corresponds to occur
check in \Fo{}.
\end{proof}

\begin{lemma} \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold
\end{lemma}
\begin{proof}[Proof sketch]
Since \elpiIn{progress1} is trivial \fstep and \hstep are the same, that is
in this
context where input terms are $\beta\eta$-normal and we disregard $\eta$-equivalence
\Ue is equivalent to \Uo.
\end{proof}

@@

\begin{lemma}
  Let $\lambda x.t$ the \rhs of a \linketa, then $\wellb{t}$.
  \label{lemma:unif-eta-aux}
\end{lemma}

\begin{proof}[Proof sketch]
  By construction, every ``problematic'' term in \Fo{} is replaced with a variable
  in the corresponding \Ho term. Therefore, $t$ is \wellb.
\end{proof}

@@

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
  \cref{prop:nf}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
  \progressetaleft, if 1) \lhs is a name, a constant or an
  application, then, $\lambda x.\lhs \appsep x$ is unified with \rhs. By
  \cref{inv:link-eta-right} and \cref{lemma:unif-eta-aux}, $\rhs = \lambda x. t$ and $\wellb(t)$.
  Otherwise, 2) \lhs has \elpiIn{lam} as functor. In both cases, unification is
  performed between terms in \wellb.
\end{proof}

@@

\begin{proof}[Proof sketch]
  \lhs is variable, and, by \cref{def:progressetaright}, \rhs is either no more
  a \maybeeta, i.e. \rhs is not a $\eta$-expansion and, so, $\wellb(\rhs)$,
  otherwise, \rhs can reduce to a term which cannot be a $\eta$-expansion, and,
  so, $\wellb(\rhs)$. In both cases, the unification between \rhs and \lhs
  is done between terms that are in \wellb.
\end{proof}

@@

\begin{proof}
  The unification is done between the \rhs of two \linketa. Both \rhs has the
  shape $\lambda x.t$, and by \cref{lemma:unif-eta-aux}, $\wellb(t)$.
  Therefore, the unification is done between well-behaved terms.
\end{proof}

@@

\begin{proof}[Proof sketch]
  By \cref{lemma:unif-eta-1,lemma:unif-eta-2,lemma:unif-eta-3}, every
  unification performed by the activation of a \linketa is done between
  terms in \wellb, therefore, the substitution remains \wellb.
\end{proof}
